---
title: "deseqOculina_orthologs"
author: "Hannah Aichelman"
date: "2/20/2024"
output: 
  html_document:
    toc: true
    theme: united
---

### Install and libray packages
```{r InstallPackages, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# This code is from Colleen Bove, lets us library the old version of DESeq (to do array quality metrics) even running versions of R after 3.6

## get Bioconductor packages
if (!requireNamespace("BiocManager"))
install.packages("BiocManager")
BiocManager::install("DESeq2")

## installing WGCNA:
#source("http://bioconductor.org/biocLite.R")
BiocManager::install(c("AnnotationDbi", "impute", "GO.db", "preprocessCore"))
BiocManager::install("WGCNA", dependencies=TRUE)
BiocManager::install("arrayQualityMetrics", dependencies=TRUE) # use this arrayQualityMetrics install if using later versions of R (3.6.3 works)
BiocManager::install("affycoretools", dependencies=TRUE)
#repos="http://cran.us.r-project.org"
## R version 3.6 is funky with arrayQualityMetrics and DESeq so need this work around:
install.packages("ape", lib = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library")
library(ape, lib.loc = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library")
install.packages("magick", lib = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library", dependencies = FALSE)
library(magick, lib.loc = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library")
BiocManager::install("arrayQualityMetrics", type = "source", checkBuilt = TRUE, lib.loc = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library")
BiocManager::install("DESeq", type = "source", checkBuilt = TRUE, lib.loc = "/Library/Frameworks/R.framework/Versions/4.0/Resources/library")
```

```{r LibraryPackages, message=FALSE, warning=FALSE}

library(DESeq2)
library(ggplot2)
library(affycoretools)
library(arrayQualityMetrics)
library(genefilter)
#library(DESeq)
library(cowplot)
library(readr)
library(RColorBrewer)
library(gplots)
library(knitr)
library(plotly)
library(vegan)
library(kableExtra)
library(reshape2)
library(prettydoc)
library(VennDiagram)
library(tidyverse)
```

###Counts files and experimental design.
```{r,cache=TRUE, echo=FALSE}
# this is the original counts file used in the analysis - which was made by mapping to a referene transcriptome of only single copy orthologs
# not using this counts file in final manuscript, follow the analyses detailed below.
#countData <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/OrthoCounts_deseq.txt") # clones are already removed from this dataset
#dim(countData)
# [1] 185  54

# New analysis - September 2023
# Try subsetting original host and symbiont count files for identified single copy orthologs
# first read in host ortholog and gene names, then combine into one df
host_ortholog_gene_names <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/O_arbuscula_ortholog_genenames.txt")
host_original_gene_names <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/O_arbuscula_original_genenames.txt")
dim(host_ortholog_gene_names) # [1] 57470     1
dim(host_original_gene_names) # [1] 57470     2

host_dat = cbind(host_original_gene_names,host_ortholog_gene_names)
colnames(host_dat) = c("host_original_gene", "host_original_gene2", "host_ortholog_gene")
host_dat = host_dat %>%
  select(host_original_gene, host_ortholog_gene)

# read in symbiont ortholog and gene names, then combine into one df
sym_ortholog_gene_names <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/B_psygmophilum_ortholog_genenames.txt")
sym_original_gene_names <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/B_psygmophilum_original_genenames.txt")
dim(sym_ortholog_gene_names) # [1] 31970     1
dim(sym_original_gene_names) # [1] 31970     2

sym_dat = cbind(sym_original_gene_names, sym_ortholog_gene_names)
colnames(sym_dat) = c("sym_original_gene", "sym_original_gene2", "sym_ortholog_gene")
sym_dat = sym_dat %>%
  select(sym_original_gene, sym_ortholog_gene)

# read in single copy ortholog names 
single_copy_orthos <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/singleCopyOrthos.txt")

# subset host gene and ortholog name dataframe to include only single copy orthologs 
host_data_sco = host_dat %>%
  filter(host_ortholog_gene %in% single_copy_orthos$V2)
# subset symbiont gene and ortholog name dataframe to include only single copy orthologs 
sym_data_sco = sym_dat %>%
  filter(sym_ortholog_gene %in% single_copy_orthos$V2)

# Need to extract only single copy orthologs from the counts files used in the host and symbiont individual gene expression analyses
# First read in the counts files for those analyses
countDataHost <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/Oculina_counts_newref_host.txt")
dim(countDataHost)
countDataSym <- read.table("~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/03_DESeq_Scripts/CountsFiles/Oculina_Counts_newref_sym.txt")
dim(countDataSym)

# subset host count data to include only single copy orthologs
countDataHost_sco = countDataHost[rownames(countDataHost) %in% host_data_sco$host_original_gene, ]
head(countDataHost_sco)
dim(countDataHost_sco)
# [1] 1578   47
# change rownames to be single copy ortholog id's instead of original gene names
rownames(countDataHost_sco) = host_data_sco$host_ortholog_gene[match(rownames(countDataHost_sco), host_data_sco$host_original_gene)]
rownames(countDataHost_sco) = single_copy_orthos$V1[match(rownames(countDataHost_sco), single_copy_orthos$V2)]

# subset symbiont count data to include only single copy orthologs
countDataSym_sco = countDataSym[rownames(countDataSym) %in% sym_data_sco$sym_original_gene, ]
head(countDataSym_sco)
dim(countDataSym_sco)
# [1] 1462   47
# change rownames to be single copy ortholog id's instead of original gene names
rownames(countDataSym_sco) = sym_data_sco$sym_ortholog_gene[match(rownames(countDataSym_sco), sym_data_sco$sym_original_gene)]
rownames(countDataSym_sco) = single_copy_orthos$V1[match(rownames(countDataSym_sco), single_copy_orthos$V2)]

# check that the host and symbiont files contain most of the same row names
rownames(countDataSym_sco) %in% rownames(countDataHost_sco)

# re-name the columns with ID, and host/symbiont 
names(countDataHost_sco) = c("OA4_C_W.host", "OA5_F_W.host", "OA6_H_W.host", 
                        "OC4_F_B.host",	"OC5_H_B.host",	"OC9_C_B.host",	
                        "OD4_C_B.host",	"OD5_F_B.host",	"OD6_H_B.host",
                        "OE10_F_W.host", "OE11_C_W.host", "OE3_H_W.host",
                        "OF7_C_B.host",	"OF8_F_B.host",	"OF9_H_B.host",
                        "OH11_F_W.host", "OH15_H_W.host", "OH1_C_W.host",
                        "OI1_C_B.host",	"OI2_F_B.host",	"OI3_H_B.host",	
                        "OJ13_C_B.host",	"OJ14_F_B.host",	"OJ15_H_B.host",	
                        "OK1_C_W.host", "OK2_F_W.host", "OK3_H_W.host",
                        "OL6_C_B.host",	"OL7_F_B.host",	"OL8_H_B.host",	
                        "OM1_C_B.host",	"OM2_F_B.host",	"OM3_H_B.host",
                        "ON4_C_W.host", "ON5_F_W.host", "ON6_H_W.host",
                        "OO1_C_W.host", "OO2_F_W.host",
                        "OP1_C_W.host", "OP2_F_W.host", "OP3_H_W.host",
                        "OQ11_H_W.host", "OQ1_C_W.host", "OQ4_F_W.host",
                        "OR7_C_B.host",	"OR8_F_B.host",	"OR9_H_B.host")

names(countDataSym_sco) = c("OA4.sym", "OA5.sym", "OA6.sym", 
                        "OC4_F_B.sym",	"OC5_H_B.sym",	"OC9_C_B.sym",	
                        "OD4_C_B.sym",	"OD5_F_B.sym",	"OD6_H_B.sym",
                        "OE10.sym", "OE11.sym", "OE3.sym",
                        "OF7_C_B.sym",	"OF8_F_B.sym",	"OF9_H_B.sym",
                        "OH11.sym", "OH15.sym", "OH1.sym",
                        "OI1_C_B.sym",	"OI2_F_B.sym",	"OI3_H_B.sym",	
                        "OJ13_C_B.sym",	"OJ14_F_B.sym",	"OJ15_H_B.sym",	
                        "OK1.sym", "OK2.sym", "OK3.sym",
                        "OL6_C_B.sym",	"OL7_F_B.sym",	"OL8_H_B.sym",	
                        "OM1_C_B.sym",	"OM2_F_B.sym",	"OM3_H_B.sym",
                        "ON4.sym", "ON5.sym", "ON6.sym",
                        "OO1.sym", "OO2.sym",
                        "OP1.sym", "OP2.sym", "OP3.sym",
                        "OQ11.sym", "OQ1.sym", "OQ4.sym",
                        "OR7_C_B.sym",	"OR8_F_B.sym",	"OR9_H_B.sym")

# this is the final merged dataframe that we will use in the DESeq model
orthoCounts = merge(countDataHost_sco, countDataSym_sco, by = 0)
dim(orthoCounts)
#[1] 1176   95

# summarize the counts per sample
orthoCounts_sum = orthoCounts %>% 
  mutate_if(is.integer, as.numeric) %>%
  select(-Row.names) %>%
  colSums()
orthoCounts_sum = as.data.frame(orthoCounts_sum)
```


Set up experimental design matrix
```{r,cache=TRUE}
# Create an experimental design matrix with updated counts file
# First remove clones and symbiont counts from aposymbiotic corals, following the same samples removed from the individual host and symbiont datasets
orthoCounts_noclones = orthoCounts %>%
  select(-c(OA4.sym, OA5.sym, OA6.sym, OE10.sym, OE11.sym, OE3.sym, OH11.sym, OH15.sym, OH1.sym, OK1.sym, OK2.sym, OK3.sym, ON4.sym, ON5.sym, ON6.sym, OO1.sym, OO2.sym, OP1.sym, OP2.sym, OP3.sym, OQ11.sym, OQ1.sym, OQ4.sym)) %>% # remove sym counts in aposymbiotic corals
  select(-c(OL6_C_B.sym,  OL7_F_B.sym,  OL8_H_B.sym)) %>% # remove clone pair
  select(-c(OP1_C_W.host, OP2_F_W.host, OP3_H_W.host, OO1_C_W.host, OO2_F_W.host, OL6_C_B.host, OL7_F_B.host, OL8_H_B.host, OK1_C_W.host, OK2_F_W.host, OK3_H_W.host, OA4_C_W.host, OA5_F_W.host, OA6_H_W.host)) %>% # remove clones and outlier genotype A from hosts
  column_to_rownames("Row.names")

# Build in identifying info from the samples
treatment = as.factor(sapply(strsplit(colnames(orthoCounts_noclones), split = "_"), "[[", 2)) %>%
  plyr::revalue(c("C" = "control", "F" = "cold", "H" = "heat"))
genotype  = as.factor(sapply(strsplit(colnames(orthoCounts_noclones), split = ""), "[[", 2))
type  = as.factor(sapply(strsplit(colnames(orthoCounts_noclones), split = "_"), "[[", 3))

expDesign = data.frame(colnames(orthoCounts_noclones), treatment, genotype, type)
expDesign$temp_type = as.factor(paste(expDesign$treatment,expDesign$type, sep = "_"))

levels(expDesign$temp_type) = c("cold_brown_host", "cold_sym_inhost", "cold_white_host",
                                "control_brown_host", "control_sym_inhost", "control_white_host",
                                "heat_brown_host", "heat_sym_inhost", "heat_white_host")

names(expDesign) = c("sample", "treatment", "genotype", "type", "treat_type")

```

Experimental Design key with sample names 
```{r,cache=TRUE, echo=FALSE}
kable(expDesign) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
Descriptive Summary stats of the mapped reads. 

```{r echo=FALSE}
stats.per.sample = data.frame(t(do.call(cbind, lapply(orthoCounts_noclones, summary))))
      stats.per.sample$libsum = apply(orthoCounts_noclones, 2, sum) ## libsum
      stats.per.sample$perc05 = apply(orthoCounts_noclones, 2, quantile, 0.05)
      stats.per.sample$perc10 = apply(orthoCounts_noclones, 2, quantile, 0.10)
      stats.per.sample$perc90 = apply(orthoCounts_noclones, 2, quantile, 0.90)
      stats.per.sample$perc95 = apply(orthoCounts_noclones, 2, quantile, 0.95)
      stats.per.sample$zeros = apply(orthoCounts_noclones==0, 2, sum)
      stats.per.sample$percent.zeros = 100*stats.per.sample$zeros/nrow(orthoCounts_noclones)
write.csv(stats.per.sample, file = "~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/08_tables/oculina_orthos_newcounts_summary_table.csv", quote = FALSE)
      
kable(stats.per.sample) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

#Outlier Detection


Conduct array quality metrics to identify outliers  
Skip while running, but arrayqualitymetrics output can be found on github

```{r, cache=TRUE}
real=newCountDataSet(countData,expDesign) 
real=estimateSizeFactors(real)
```

```{r, fig.align='center', cache=TRUE }
plot(sort(sizeFactors(real)), main = "Ocuina Ortholog Size Factors") 
```
<br>
Size factors of each sample used to normalize between libraries. 
<br> 
<br> 
Create a directory with a bunch of output figures from arrayQualityMetrics() for you to explore if their are outliers. 

```{r message=FALSE, warning=FALSE, eval = FALSE}
cds=estimateDispersions(real,method="blind", fitType="local")
vsdBlind=varianceStabilizingTransformation(cds)
arrayQualityMetrics(vsdBlind,intgroup=c("Type_Temp"), force=TRUE, outdir = "Oculina_ortholog_arrayqualitymetrics") # this makes a directory "arrayQualityMetrics" and outputs various reports on outliers
```

Look at total counts across Oculina host and symbiont samples
```{r}
orthoCounts = sapply(orthoCounts_noclones, as.numeric)
totalCounts=colSums(orthoCounts)
barplot(totalCounts, ylab="raw counts", main = "Oculina ortholog total counts", las=2)

min(totalCounts) 
max(totalCounts) 

```
###Outlier Conclusisons

Do not need to remove any outliers for the Oculina host or symbiont. 


Look at total counts across symbiont samples in the host and in culture
```{r, fig.align='center', cache=TRUE}
orthoCounts = sapply(orthoCounts_noclones, as.numeric)
totalCounts=colSums(orthoCounts)
barplot(totalCounts, ylab="raw counts", main = "Oculina ortholog total counts", las=2)

min(totalCounts) 
max(totalCounts) 

```


#Differential Expression


```{r warning=FALSE, cache=TRUE}
dds<-DESeqDataSetFromMatrix(countData=orthoCounts_noclones, colData=expDesign, design=~treat_type) #can only test for the main effects of treatment
keep.dds = rowSums(counts(dds) >= 2) >= 42 # at least 33 samples (80%) with a count of 2 or higher
dds.corr = dds[keep.dds,]

dds = DESeq2::DESeq(dds.corr)
results = results(dds) 
summary(results) # 134 up 155 down / 23 up 47 down after filtering
plotDispEsts(dds)
```


Let's check to see if we set up our contrast correctly. We should have the treatment condition first and the control second in the log2 fold change (MLE) output. 

```{r, cache=TRUE}
head(results)
resultsNames(dds)
```


```{r echo=FALSE, cache=TRUE}
norm.counts.ortho.new = counts(dds, normalized = TRUE) # these are the counts DESeq uses
write.csv(norm.counts.ortho, "~/Dropbox/BU/Host_Buffering/FinalGithub/Oculina_Host_Sym_GE/08_tables/Oculina_normalized_counts_ortho.csv") #these are all counts, not considering treatment comparisons
              
norm.counts.stats = data.frame(
  min = apply(norm.counts.ortho, 2, min),
  mean = apply(norm.counts.ortho, 2, mean),
  median = apply(norm.counts.ortho, 2, median), 
  max = apply(norm.counts.ortho, 2, max),
  zeros = apply(norm.counts.ortho == 0, 2, sum), 
  percent.zeros = 100* apply(norm.counts.ortho == 0, 2, sum) / nrow(norm.counts.ortho) 
)

kable(norm.counts.stats) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

# do a bar plot to make sure there is no consistent difference in counts by sample type
norm.counts.sum.new = colSums(norm.counts.ortho.new)
barplot(norm.counts.sum.new, col=expDesign$treat_type, ylab="counts", main = "Oculina Normalized Counts - New Counts")

```

Lets do a rlogged and vst transformation, which is useful various unsupervised clustering analyses. Be sure the include blind = TRUE as it doesn't normalize the data with any priors from our experimental design. 

```{r, cache=TRUE}
rlogged = rlogTransformation(dds, blind = TRUE)
vst = varianceStabilizingTransformation(dds, blind = TRUE)
```

#Expression Comparisons

Hot versus control expression comparison - Brown Host
```{r Brown Host Hot vs control}
##second term is the "control"
resBrownHostH.BrownHostC=results(dds, contrast=c("treat_type","heat_brown_host","control_brown_host"), independentFiltering = F)
head(resBrownHostH.BrownHostC)
table(resBrownHostH.BrownHostC$pvalue<0.05)

table(resBrownHostH.BrownHostC$padj<0.1)
summary(resBrownHostH.BrownHostC)
#write.table(resBrownHostH.BrownHostC, file="/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/brown_host_hot_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resBrownHostH.BrownHostC, main = "Hot vs Control, Brown Host Ortholog")
```


Cold versus control expression comparison - Brown Host
```{r Brown Host Freezing vs Control}
#compare host freezing to host control
resBrownHostF.BrownHostC=results(dds, contrast=c("treat_type","cold_brown_host","control_brown_host"), independentFiltering = F)
head(resBrownHostF.BrownHostC)
table(resBrownHostF.BrownHostC$pvalue<0.05)

table(resBrownHostF.BrownHostC$padj<0.01)
summary(resBrownHostF.BrownHostC)

#write.table(resBrownHostF.BrownHostC, "/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/brown_host_cold_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resBrownHostF.BrownHostC, main = "Freezing vs Control, Brown Host Orthologs")
```

Hot versus control expression comparison - White Host
```{r White Host Hot vs Control}
##second term is the "control"
resWhiteHostH.WhiteHostC=results(dds, contrast=c("treat_type","heat_white_host","control_white_host"), independentFiltering = F)
head(resWhiteHostH.WhiteHostC)
table(resWhiteHostH.WhiteHostC$pvalue<0.05)

table(resWhiteHostH.WhiteHostC$padj<0.1)
summary(resWhiteHostH.WhiteHostC)
#write.table(resWhiteHostH.WhiteHostC, file="/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/white_host_hot_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resWhiteHostH.WhiteHostC, main = "Hot vs Control, White Host Ortholog")
```

Cold versus control expression comparison - White Host
```{r  White Host Freezing vs Control}
#compare host freezing to host control
resWhiteHostF.WhiteHostC=results(dds, contrast=c("treat_type","cold_white_host","control_white_host"), independentFiltering = F)
head(resWhiteHostF.WhiteHostC)
table(resWhiteHostF.WhiteHostC$pvalue<0.05)

table(resWhiteHostF.WhiteHostC$padj<0.01)

summary(resWhiteHostF.WhiteHostC)

#write.table(resWhiteHostF.WhiteHostC, "/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/white_host_cold_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resWhiteHostF.WhiteHostC, main = "Freezing vs Control, Brown Host Orthologs")
```

Hot versus control expression comparison - Symbiont In Host
```{r Sym in Host Hot vs Control}
##second term is the "control"
resSymInHostH.SymInHostC=results(dds, contrast=c("treat_type","heat_sym_inhost","control_sym_inhost"), independentFiltering = F)
head(resSymInHostH.SymInHostC)
table(resSymInHostH.SymInHostC$pvalue<0.05)


table(resSymInHostH.SymInHostC$padj<0.05)

summary(resSymInHostH.SymInHostC)
#write.table(resSymInHostH.SymInHostC, file="/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/sym_inhost_hot_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resSymInHostH.SymInHostC, main = "Hot vs Control, Sym In Host Ortholog")
```
Cold versus control expression comparison - Symbiont In Host
```{r Sym in Host Freezing vs Control}
##second term is the "control"
resSymInHostF.SymInHostC=results(dds, contrast=c("treat_type","cold_sym_inhost","control_sym_inhost"), independentFiltering = F)
head(resSymInHostF.SymInHostC)
table(resSymInHostF.SymInHostC$pvalue<0.05)

table(resSymInHostF.SymInHostC$padj<0.05)

summary(resSymInHostF.SymInHostC)
#write.table(resSymInHostF.SymInHostC, file="/Users/hannahaichelman/Documents/BU/Host_Buffering/MPCC_2018/Sym_analyses/tables/sym_inhost_cold_orthos_results.txt", quote=F, sep="\t")
```

MA plot, red dots indicate genes significantly differentially expressed at FDR < 0.1 

```{r, cache=TRUE, fig.align='center'}
DESeq2::plotMA(resSymInHostF.SymInHostC, main = "Freezing vs Control, Sym In Host Ortholog")
```


## Make a barplot of differentially expressed genes
```{r, cache=TRUE, fig.align='center'}
summary(resBrownHostH.BrownHostC)
# up new counts filtered = 2
# down new counts filtered = 10
summary(resBrownHostF.BrownHostC)
# up new counts filtered = 40
# down new counts filtered = 21
summary(resWhiteHostH.WhiteHostC)
# up new counts filtered = 5
# down new counts filtered = 3
summary(resWhiteHostF.WhiteHostC)
# up new counts filtered = 28
# down new counts filtered = 14
summary(resSymInHostH.SymInHostC)
# up new counts filtered = 1
# down new counts filtered = 3
summary(resSymInHostF.SymInHostC)
# up new counts filtered = 1
# down new counts filtered = 2

ortho_combined_degs = data.frame(
  temperature = factor(c("Heat","Heat","Cold", "Cold",
                         "Heat","Heat","Cold", "Cold",
                         "Heat","Heat","Cold", "Cold")),
  type = factor(c("Sym_Host", "Sym_Host", "Sym_Host","Sym_Host",
                  "Apo_Host","Apo_Host","Apo_Host","Apo_Host",
                  "Sym_InHost", "Sym_InHost","Sym_InHost", "Sym_InHost")),
  direction = factor(c("up","down","up","down",
                       "up","down","up","down",
                       "up","down","up","down")),
  type_temperature_direction = factor(c("sym_host_heat_up","sym_host_heat_down",
                                        "sym_host_cold_up","sym_host_cold_down",
                                        "apo_host_heat_up","apo_host_heat_down",
                                        "apo_host_cold_up","apo_host_cold_down",
                                        "sym_inhost_heat_up","sym_inhost_heat_down",
                                        "sym_inhost_cold_up","sym_inhost_cold_down")),
  number = c(2, -10, 40, -21, 
             5, -3, 28, -14,
             1, -3, 1, -2)
)

annotate_text = data.frame(
  label = c("4", "14", "34", "22","5", "2","31", "11","1","6","3","2"),
  temperature = factor(c("Heat","Heat","Cold", "Cold",
                         "Heat","Heat","Cold", "Cold",
                         "Heat","Heat","Cold", "Cold")),
  type = factor(c("Sym_Host", "Sym_Host", "Sym_Host","Sym_Host",
                  "Apo_Host","Apo_Host","Apo_Host","Apo_Host",
                  "Sym_InHost", "Sym_InHost","Sym_InHost", "Sym_InHost")),
  direction = factor(c("up","down","up","down",
                       "up","down","up","down",
                       "up","down","up","down")),
  type_temperature_direction = factor(c("sym_host_heat_up","sym_host_heat_down",
                                        "sym_host_cold_up","sym_host_cold_down",
                                        "apo_host_heat_up","apo_host_heat_down",
                                        "apo_host_cold_up","apo_host_cold_down",
                                        "sym_inhost_heat_up","sym_inhost_heat_down",
                                        "sym_inhost_cold_up","sym_inhost_cold_down")),
  x = c(1,1,1,1,1,1,2,2,2,2,2,2),
  y = c(20,-20,20,-20,20,-20,20,-20,20,-20,20,-20)
)


deg_cols = c("sym_host_cold_up" = "#3182bd", "sym_host_cold_down" =  "#9ecae1","sym_host_heat_up" = "#de2d26", "sym_host_heat_down" ="#fc9272", "apo_host_cold_up" = "#3182bd", "apo_host_cold_down" =  "#9ecae1","apo_host_heat_up" = "#de2d26", "apo_host_heat_down" ="#fc9272","sym_inhost_cold_up" = "#74c476", "sym_inhost_cold_down" =  "#c7e9c0","sym_inhost_heat_up" = "#fd8d3c", "sym_inhost_heat_down" ="#fdd0a2")

deg_barplot = ggplot(ortho_combined_degs, aes(x = temperature, y = number, fill=type_temperature_direction)) +
  geom_bar(stat = "identity", position = position_stack(), color = "black") +
  scale_fill_manual(values = deg_cols) +
  xlab("Temperature Treatment") +
  ylab("Differentially expressed orthologs") +
  theme_bw() +
  theme(legend.position = "none") +
  facet_grid(. ~ type) +
  theme(panel.spacing = unit(0, "lines")) #+
  #geom_text(data = annotate_text, aes(x = x, y = y, label = label))
deg_barplot

ggsave(deg_barplot, file = "orthos_deg_barplot_newcounts_filtered.pdf", width=5, height=4, units=c("in"), useDingbats=FALSE)
```

##Test for differences in DE orthologs
```{r, cache=TRUE, fig.align='center'}
summary(resBrownHostH.BrownHostC)
# up new counts filtered = 2
# down new counts filtered = 10
summary(resBrownHostF.BrownHostC)
# up new counts filtered = 40
# down new counts filtered = 21
summary(resWhiteHostH.WhiteHostC)
# up new counts filtered = 5
# down new counts filtered = 3
summary(resWhiteHostF.WhiteHostC)
# up new counts filtered = 28
# down new counts filtered = 14
summary(resSymInHostH.SymInHostC)
# up new counts filtered = 1
# down new counts filtered = 3
summary(resSymInHostF.SymInHostC)
# up new counts filtered = 1
# down new counts filtered = 2

# Out of 152 nonzero total read counts for new counts file filtered to base mean of 2
res.brownhost_v_whitehost_heat <- prop.test(x = c(12, 8), n = c(152, 152), alternative = 'greater')

res.brownhost_v_whitehost_heat 
# data:  c(12, 8) out of c(152, 152)
# X-squared = 0.48169, df = 1, p-value = 0.2438
# alternative hypothesis: greater
# 95 percent confidence interval:
#  -0.02697301  1.00000000
# sample estimates:
#     prop 1     prop 2 
# 0.07894737 0.05263158 

res.brownhost_v_whitehost_cold <- prop.test(x = c(61,42), n = c(152, 152), alternative = 'greater')

res.brownhost_v_whitehost_cold 
# data:  c(61, 42) out of c(152, 152)
# X-squared = 4.7576, df = 1, p-value = 0.01458
# alternative hypothesis: greater
# 95 percent confidence interval:
#  0.02990058 1.00000000
# sample estimates:
#    prop 1    prop 2 
# 0.4013158 0.2763158 

res.brownhost_v_syminhost_heat <- prop.test(x = c(12, 4), n = c(152, 152), alternative = 'greater')

res.brownhost_v_syminhost_heat 
# data:  c(12, 4) out of c(152, 152)
# X-squared = 3.2326, df = 1, p-value = 0.03609
# alternative hypothesis: greater
# 95 percent confidence interval:
#  0.004215101 1.000000000
# sample estimates:
#     prop 1     prop 2 
# 0.07894737 0.02631579 

res.brownhost_v_syminhost_cold <- prop.test(x = c(61, 3), n = c(152, 152), alternative = 'greater')

res.brownhost_v_syminhost_cold
# data:  c(61, 3) out of c(152, 152)
# X-squared = 64.303, df = 1, p-value = 5.334e-16
# alternative hypothesis: greater
# 95 percent confidence interval:
#  0.3070225 1.0000000
# sample estimates:
#     prop 1     prop 2 
# 0.40131579 0.01973684 

res.whitehost_v_syminhost_heat <- prop.test(x = c(8, 4), n = c(152, 152), alternative = 'greater')

res.whitehost_v_syminhost_heat 
# data:  c(8, 4) out of c(152, 152)
# X-squared = 0.78082, df = 1, p-value = 0.1884
# alternative hypothesis: greater
# 95 percent confidence interval:
#  -0.01691832  1.00000000
# sample estimates:
#     prop 1     prop 2 
# 0.05263158 0.02631579 

res.whitehost_v_syminhost_cold <- prop.test(x = c(42, 3), n = c(152, 152), alternative = 'greater')

res.whitehost_v_syminhost_cold
# data:  c(42, 3) out of c(152, 152)
# X-squared = 37.664, df = 1, p-value = 4.202e-10
# alternative hypothesis: greater
# 95 percent confidence interval:
#  0.1875205 1.0000000
# sample estimates:
#     prop 1     prop 2 
# 0.27631579 0.01973684 

```

##Principal Component Analyses 
Not using PCA for the ortholog analysis anymore

<br>
First we create a PCA data frame and calculate the variance estimated by PC1 and PC2
```{r cache=TRUE, fig.align="center"}
pcadata = DESeq2::plotPCA(rlogged, intgroup = "treat_type", returnData = TRUE)
percentVar = round(100 * attr(pcadata, "percentVar"))
pca = prcomp(t(assay(rlogged)), center = TRUE, scale. = FALSE)
```
<br><br>

Using adonis from library(vegan) we can see if there are any significant differences based on treatment (as is illustrated in the model displayed below) 
```{r cache=TRUE}
adonis2(pca$x ~ treat_type, data = pcadata, method = 'eu')
```
<br><br>
Significant effect of treatment (p = 0.001)
```{r cache=TRUE, fig.align="center"}
cols = c("control_brown_host" = "grey", "cold_brown_host" = "#3182bd", "heat_brown_host" =  "#de2d26",
         "control_white_host" = "grey", "cold_white_host" = "#3182bd", "heat_white_host" = "#de2d26",
         "control_sym_inhost" = "#a6611a", "cold_sym_inhost" = "#74c476", "heat_sym_inhost" = "#fd8d3c",
         "control_culture" = "#a6611a", "cold_culture" = "#74c476", "heat_culture" = "#fd8d3c")
         

shapes = c("control_brown_host" = 19, "cold_brown_host" = 19, "heat_brown_host" =  19,
           "control_white_host" = 21, "cold_white_host" = 21, "heat_white_host" = 21,
           "control_sym_inhost" = 19, "cold_sym_inhost" = 19, "heat_sym_inhost" = 19,
           "control_culture" = 21, "cold_culture" = 21, "heat_culture" = 21)

legend.labels =  c("control_sym_host", "cold_sym_host", "heat_sym_host",
           "control_apo_host", "cold_apo_host", "heat_apo_host",
           "control_sym_inhost", "cold_sym_inhost", "heat_sym_inhost",
           "control_culture", "cold_culture", "heat_culture")

plotPCA_rlog_batchcorr = DESeq2::plotPCA(rlogged, returnData = TRUE, intgroup = "treat_type" ) %>% 
      ggplot(aes(x = PC1, y = PC2)) +
      geom_point(aes(colour = treat_type, shape = treat_type), size = 3) +
      stat_ellipse(aes(x = PC1, y = PC2, color=treat_type, lty = treat_type), 
                   type = "t", 
                   lwd = 1)+    
      scale_shape_manual(values = shapes,
                         name = "Treatment",
                         labels = legend.labels) +
      scale_colour_manual(values = cols,
                          name = "Treatment",
                          labels = legend.labels) +
      scale_fill_manual(values = cols,
                        name = "Treatment",
                        labels = legend.labels) +
      stat_ellipse(aes(color=rlogged$treat_type, lty = rlogged$treat_type), type = "t", lwd = 1) +
      scale_linetype_manual(values = c(1,1,1,2,2,2,1,1,1,2,2,2),
                            name = "Treatment",
                            labels = legend.labels) +
      xlab(paste0("PC1: ",percentVar[1],"% variance")) +
      ylab(paste0("PC2: ",percentVar[2],"% variance")) +
      #labs(title = "~ Type_Temp, p<0.001")+
      theme_bw()
plotPCA_rlog_batchcorr

ggsave(plotPCA_rlog_batchcorr, filename = "orthos_pca.pdf", width=7, height=5, units=c("in"), useDingbats=FALSE)
```
Principal component analyses highlighting treatment groups. Treatment has a significant affect (p < 0.001) on overall expression




